Stochastic blockmodeling of relational event dynamics

### 2012-05-31 docs/intro.tex
Edited the intro section, and added a reference.  (BTW, did you really intend W+F to be 1992?  You seem to have been citing their book proposal, instead of the 1994 book itself (albeit not in this paper).)

Statistical methods for analyzing network data are becoming increasingly useful for studying phenomena ranging from online social behavior to protein  interactions \cite{Goldenberg2009}.
We often  observe events occurring between nodes over time, and a common goal is to understand the differences in the total number of interactions --- perhaps with respect to information known about the nodes.
%In this case, we have a count for the total activity of each dyad, and one may analyze it has a static network. 
%Methods for static networks such as a 
Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of techniques which can model this heterogeneity by assuming latent subgroups of nodes have similar tendencies in their group-wise rate of interaction.

Network-based event data are powerful, however, because we enable the study of the processes governing these events.
Theories about the dynamics among nodes will be often be difficult to study when only using the marginal rate of events as illustrated in Figure \ref{fig:introexample}
For example, a common aspect of human communication is \emph{reciprocity} --- after person A speaks to person B, B subsequently has a higher propensity to speak to A.  
These sorts of dynamics may be studied with event data, but again, our focus may be on the differences in these dynamics among the nodes in the network.  
In the context of studying interactions among students, one might be interested particular conversational tendencies are associated with covariates about the student or the classroom.  
A standard blockmodel only provides a means for studying the structure of marginal rates of communication, but not the shared tendencies in the dynamics of the sequence of events.

@@

\begin{SCfigure}
\centering
\includegraphics[scale=.45]{../figs/introexample/all}
\caption{Illustration of differentiation by rate of interaction versus by dynamics.  The left column shows an example where within-block communication is large, and there is a higher tendency for reciprocity within a block.  In the situation where the two groups are not differentiated by rate, as in the right column, a standard blockmodel is unable to distinguish between groups A and B.  The proposed method can, however, learn such groups by employing more flexible definitions of shared dynamics.}
\label{fig:example}
\end{SCfigure}

>>>>

Statistical methods for analyzing network data are becoming increasingly useful for studying phenomena ranging from online social behavior to protein  interactions \cite{Goldenberg2009}.
Recent work has expanded to include settings in which we observe events occurring between nodes over time (i.e., \emph{relational events}, as opposed to static edge structures or ongoing relationships), with the common goal of modeling interaction dynamics in terms of both endogenous mechanisms and exogenous covariates.  
A key concern in this regard is the ability to detect differential behavioral tendencies on the part of subsets of nodes, the dynamic analog of \emph{role structure} in classical network analysis \cite{Wasserman1994}.

In the cross-sectional case, \emph{stochastic blockmodels} \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} have been proposed as a family of approaches that capture behavioral similarity by identifying subsets of nodes with similar patterns of ties to those in other sets.  While it is natural to apply these ideas directly to relational event data via blockmodeling of the time-marginalized rates of interaction among dyads (effectively treating the event structure as a valued graph), there are limits to what this approach can detect.  Consider, e.g., Figure~\ref{fig:introexample}.  In the top left panel, we depict the time-marginalized frequencies of simulated interactions between members of two groups (A and B), with darker cells indicating higher interaction frequencies.  As can be seen, a clear role structure is present, with members of each subgroup interacting at higher rates with co-members than out-group members; such a structure is easily detectable via conventional blockmodeling techniques.  By contrast, consider the interaction patterns shown in the top right panel.  Here, there is no systematic difference in marginal rates between the two groups, rendering them invisible under a standard blockmodeling approach.

There is, however, a difference to be detected.  The bottom panels of Figure~\ref{}fig:introexample} show, for each respective simulation, parameters (as discussed in Section~\ref{}) governing the tendency towards reciprocation for members of each group vis a vis communications coming from in- or out-group members.  For both simulated cases, the log-hazard for an event in which a member of group A or group B immediately responds to an incoming communication from an in-group member is increased by an increment of 2 (relative to a non-reciprocating event), and the log-hazard for an event in which a member of group B immediately responds to an incoming communication from a member of group A is increased by a corresponding increment of 0.5.  Groups A and B are thus distinctive in terms of their dynamic behavior, even in the absence of marginal differences in propensity to communicate.  A blockmodeling approach that classifies nodes based on shared dynamics (rather than merely shared marginal communication rates) can potentially identify such subtle distinctions; in this paper, we introduce such an approach.

%For example, a common aspect of human communication is \emph{reciprocity}: person A speaks to person B, then B has a higher propensity to speak to A.  
%Through an event-based analysis we can study how nodes vary in their tendencies to behave in this manner.  %the similarity --- and differences --- in the dynamics of nodes in the network. 
%This is particularly relevant when studying interactions among students in a classroom, where differences in conversational tendencies might be associated with covariates about the student or classroom, or based on unobserved groupings, and may have ramifications for education.
%In such contexts, a standard blockmodel only provides a means for studying the structure of marginal rates of communication, but not the shared tendencies in the dynamics of the sequence of events.

@@

\begin{SCfigure}
\centering
\includegraphics[scale=.45]{../figs/introexample/all}
\caption{Illustration of differentiation by rate of interaction versus by dynamic behavior.  The left column shows an example where within-block communication is large, and there is a higher tendency for reciprocity within a block.  In the situation where the two groups are not differentiated by rate, as in the right column, a standard blockmodel is unable to distinguish between groups A and B.  The proposed method can, however, learn such groups by employing more flexible definitions of shared dynamics. {\color{red} (Suggest renaming "Dynamics parameters" to "Behavioral parameters" or something.  "Dynamics parameters" is not gramatically sound.)}}
\label{fig:example}
\end{SCfigure}
